---
title: "R Notebook"
output:
  pdf_document: default
  html_notebook: default
---

## ggplot workshop

#### Watzek Library
#### January 15th, 2020


### Goals of this workshop:  

* Load and filter datasets as needed
* Learn the basic elements of ggplot
* 


### What is ggplot?

ggplot is an R package created by Hadley Wickham (he's kind of a big deal in the R world). The "gg" translates to "grammar of graphics", and is founded in the idea that all data visualizations are comprised of three components:

* data set
* aesthetics, or visual marks that represent the data (i.e. the stuff that you see)
* geometric objects, or "geoms" (e.g the type of plot)

ggplot is part of the "tidyverse" - a series of packages that share common strategies for working with data and generating visualizations. In last Wednesday's workshop, we used some tidyverse packages (readr, dplyr) for reading and filtering datasets. Fortunately, there's a simple way to load all these libraries at once, as we'll use various functions spread throughout different packages in the tidyverse:

```{r}
library(tidyverse)
```


### Getting Help  
ggplot has many, many options and possibilities. When getting started (and even once you're a ggplot expert), it's a good idea to refer to documentation. In the "Files" section, click on the "ggplot2-cheatsheet.pdf". The RStudio "help" function is also useful:

```{r}
help(ggplot)
```

```{r}
help(geom_jitter)
```

As always, entering your question or error message into your favorite search engine is trypically a pretty effective way to get help. Chances are good that someone else has already encountered a ggplot problem you've encountered, and the answers are likely online.

### Sample plots

Load data set
```{r}
babyNames<-read_csv("babyNames.csv")

#prints babyNames
babyNames
```



Let's say were interested in how the number of female babies annually named Jane has changed over time. First we can use the "filter" function to get a subset of our data:
```{r}

jane<-filter(.data=babyNames, (Name=="Jane" & Sex=="Girls"))

# print "jane" dataset
jane

```


Our "jane" variable now just has the count of "Janes" over time. Let's do a line plot to see the change over time. Here is the code we will run:

ggplot(data=jane, mapping=aes(x=Year, y=Count))+geom_line()

Looking at the code, here's how we're including the three basic parts of a ggplot chart: 

* DATA - data=jane  
* AESTHETICS - mapping=aes(x=Year, y=Count)  
* GEOM - +geom_line()


```{r}

ggplot(data=jane, mapping=aes(x=Year, y=Count))+geom_line()


```


YOUR TURN:
First, create a dataset variable limited to just your name. Then, generate a line plot of your name. 
```{r}


```

Bonus! Can you find and plot a name for which the use is increasing over time?
```{r}



```



Area Plots!

We can change our plot type:

```{r}

ggplot(data=jane, aes(x=Year, y=Count))+geom_area()

```


YOUR TURN:
Create an area plot of your name over time
```{r}


```



### Bells and Whistles

You can specify labels to add to your plot by using the "labs()" function:

```{r}

ggplot(data=jane, aes(x=Year, y=Count))+geom_area()+labs(y="number of Janes", x="Over Time", title="Janes over time")

```

Additionally you can add colors to the area:

```{r}
ggplot(data=jane, aes(x=Year, y=Count))+geom_area(color = "black", fill = "blue")+labs(y="number of Janes", x="Over Time", title="Janes over time in blue!")


```


YOUR TURN:
Can you add some labels and color to your area plot?
```{r}


```




### More data filtering with the "pipe" operator and summarise()

Let's take a look at a built-in dataset called "chickwts":

```{r}

chickens<-as_tibble(chickwts)
# as_tibble() puts the dataset in "tibble" format

chickens
```

This data is from a study in 1948 comparing the weights of chickens on different diets. Let's say we're interested in comparing the average weights of the chickens grouped by the type of diet, with the end goal of displaying the data in a bar chart. We can readily get our data in the format we need by using using the "pipe" operator, combined with group_by() and summarise(): 

```{r}
averageChickens<-chickens %>% group_by(feed) %>% summarise(avg=mean(weight))

averageChickens
```

What is happening here? The "%>%" (or "pipe") operator sort of means "pass this on to the next thing". It's a handy R convention for chaining functions together.  

The "summarise" (a.k.a. "summarize") function, when combined with "group_by", is good for adding new statistical metrics for the groups. Like the function "mutate()", summarise creates a column - we're naming it "avg".

Now we can take this data "averageChickens" and create a bar chart:

```{r}

ggplot(data=averageChickens, mapping=aes(x=feed, y=avg))+geom_bar(stat="identity")

# stat="identity" means "the y value is part of the dataset"

# another common stat value is "count", where the height of the bar is based upon counting the data rows:
# ggplot(data=chickwts, mapping=aes(x=feed))+geom_bar(stat="count")

```

This seems fine, but we can make it a little better by adding some labels, and using the "fill" aes attribute to add some color and a legend:

```{r}
ggplot(data=averageChickens, mapping=aes(x=feed, y=avg, fill=feed))+geom_bar(stat="identity")+labs(title="Average Chicken weight by Food Type", x="Food Type", y="Units of Weight", fill = "Food Type")

# note that "fill" is an option in labs() - this gives a title to the legend

```


### "Final Project"

It's now your turn to make a data visualization! 


Some options include (but are not limited to):  

* Use the babyNames dataset to compare two names over time (hint: use the "or" operator to filter the dataset)
* Use the PlantGrowth dataset to create a bar chart comparing the different weights of plant growth
* Try a new chart type (refer to the ggplot cheat sheet)
* Make a chart from a dataset that interests you (in the console, run data()  to view the available datasets)

Suggested strategy:

* Find/examine a dataset, and decide upon a useful plot from that data  
* Filter your data to get a subset of the larger dataset (if applicable). Consider using group_by() with summarise() if you want to plot statistical metrics (mean, standard deviation, etc.)  
* Create an intial plot with the three basic components (data, aesthetics, geom).
* Augment your plot by adding labels, colors, etc.


YOUR TURN:
Make a graph of your choosing, maybe share w/class!
```{r}






```




### Saving your plot 

To save your plot as a file, you can use the ggsave() function. ggsave will save the last plot generated:

```{r}

ggsave("plot.png", height=5, width=5)

```

From there, you can download your pot file, and use it however you choose.



Parting Thoughts:

* Learning any language takes persistence and effort. 
* Pick a project or dataset relevant to your work, and try using it in R.
* Need some local help? Reach out to us - digitalinitiatives@lclark.edu 


Online tutorials:

* http://www.codeschool.com/courses/try-r
* http://www.datacamp.org
* http://swirlstats.com/
* http://r4ds.had.co.nz/  

Getting help:

* Documentation and tutorials: http://cran.r-project.org/other-docs.html
* Recommended R packages by topic: http://cran.r-project.org/web/views/
* Mailing list: https://stat.ethz.ch/mailman/listinfo/r-help
* StackOverflow: http://stackoverflow.com/questions/tagged/r
* Just google your question (e.g. "How do I do 'x' in R?")

